---
title: "Redlining and Health Outcomes in Jackson, MS"
date: "Sarah Bolongaita, Hae In Kim, Christina Scaduto, Lily Shay"
output: 
  html_document: 
    highlight: tango
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Overview and Motivation
Too often historical policies are dismissed as unimportant and no longer having an impact in the present day. Time and again, research and evidence have illustrated this to be untrue and we want to demonstrate this using data science tools obtained in this course. 

Our project examines the relationship between redlining in Jackson, MS and various health indicators and outcomes. As a part of FDR’s New Deal, a federal agency known as the Home Owners’ Loan Corporation implemented the redlining practice in many cities across the United States. Neighborhoods were “evaluated” based on their “mortgage security” and areas that posed the lowest risks were graded “A” or “Best.” Neighborhoods with the “highest risk” were graded “D” or “Hazardous.” These evaluations were racist and inflicted even more structural violence in communities of color. 

Under redlining it was impossible for people of color to receive mortgage loans to purchase homes in A/B graded areas (“Best,” “Still Desirable”) neighborhoods. In the US, owning property is considered the most significant way to generate wealth and hold intergenerational wealth and redlining prevented this for many families of color across the US. Evaluating areas with low property value also deflected investment in public goods, such as public schools that are typically paid for by property taxes. Additionally, businesses were also unable to receive insurance in C/D graded neighborhoods, perpetuating divestment and decreasing access to resources (i.e., health care systems, grocery stores, etc.).

In short, redlining was government-sanctioned segregation. 

While the Fair Housing Act of 1968 made formal redlining illegal, it did not create a mechanism to undo past harms. The lack of investment in and divestment from these communities has largely gone uncorrected and continues to shape Jackson. 

By comparing the HOLC grades that were assigned in the 1930s to health data from the past 10 years, we hope to illuminate that the harm redlining imposed on these neighborhoods is not a thing of the past, but lingers to the present day. 

# Related Work 
The project stemmed from preliminary research that a group member, Lily Shay, conducted for a summer practicum. Her practicum focused specifically on the impact of redlining in Jackson, MS on food insecurity and relevant health outcomes. 

This project builds on Lily's project by expanding the data set to include multi-year data and explore other the relationship between redlining and other health related measures. 

## How questions evolved
Our group initially ran into challenges pulling multi-year data from the City Health Dashboard. We turned to the 500 Cities Project data - the source of data that the Dashboard pulls from. 

In addition to this, we had trouble parsing out demographic data related to the census tracts that we were looking at. Because of this, we let go of our questions regarding the differential effect of redlining based on race/ethnicity and area median income.

After compiling multi-year data (2013-2017), we saw a list of 28 specific health measures available for the census tracts in Jackson, MS. This list informed what issues we wanted to explore. For example, we saw there was data available for diabetes, high blood pressure and obesity, but there were no indicators for life expectancy or mental distress. 

We conducted exploratory analysis on these measures and their relationship to HOLC grades. We found that most of the associations were significant. With this in mind, we decided to focus on current lack of insurance, obesity, diabetes, and mental health. 

# Data 
## Data Sources
** put links**
Data was obtained from City Health Dashboard, the CDC's 500 Cities Project Data. The 500 Cities Project is a collaboration between the Robert Wood Johnson Foundation and the CDC Foundation. The 500 Cities Project provides city and census tract-level small area estimates for health outcomes. 

## Data Cleanup
Health measure data was acquired at a census tract-level for Jackson, MS from the 500 Cities Projects Data. Data from 4 different years were compiled into one data frame. Mismatched names between data frames were checked and addressed. Data frames were sorted by year, measure and ID.

For all graphs, HOLC grades are interpreted as follows: 
A: 'Best' 
B: 'Still Desirable'
C: 'Definitely Declining' 
D: 'Hazardous'  

Since HOLC graded areas do not perfectly coincide with census tracts, initial spatial analysis was done in ArcGIS to establish what percentage of the HOLC graded areas were within certain census tracts. 

In our analysis, census tracts and corresponding data were treated as one of these grades if 20% or more of the land was graded a particular grade. For example, if census tract 401 encompassed 40% of a Grade A area, it coded as Grade A data. “No grade” areas either were not designated a grade in the 1930s or had less than 20% of the census tract area had been graded. These areas will be used as our ‘control’ group. 

Looking through the list of 28 measures, we realized that not all measures were collected every year. We chose 4 to focus on - health insurance, obesity, diabetes, and mental health. These indicators were collected in the same years - 2014, 2015, and 2017.

### Loading Packages
```{r}
# Creating a function for loading packages
ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)) 
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE)
}

# Loading packages
packages <- c("dplyr", "DT", "foreign", "janitor", "maditr", "readr", "readxl", "reshape", "stringr", "tidyr")
ipak(packages)

# Clearing environment
rm(list = ls())

# Setting working directory
```

<br>
<br>

### Health Data

#### Compiling datasets published from 2016-2019

```{r}
# Loading data for available years
df.2016 <- read_csv("500_Cities__Local_Data_for_Better_Health__2016_release.csv")
df.2017 <- read_csv("500_Cities__Local_Data_for_Better_Health__2017_release.csv")
df.2018 <- read_csv("500_Cities__Local_Data_for_Better_Health__2018_release.csv")
df.2019 <- read_csv("500_Cities__Local_Data_for_Better_Health__2019_release.csv")

# Comparing dfs
comparison <- compare_df_cols(df.2016, df.2017, df.2018, df.2019)

# Fixing mismatched names
df.2016 <- df.2016 %>% dplyr::rename("PopulationCount" = Population2010)
df.2017 <- df.2017 %>% dplyr::rename("PopulationCount" = Population2010)
df.2018 <- df.2018 %>% dplyr::rename("GeoLocation" = Geolocation)

# Saving original variable order 
original <- names(df.2016)

# Sorting variable order alphabetically
sorted <- sort(names(df.2016))

# Sorting all df columns alphabetically in preparation of rbind
df_2016 <- df.2016 %>% select(sorted)
df_2017 <- df.2017 %>% select(sorted)
df_2018 <- df.2018 %>% select(sorted)
df_2019 <- df.2019 %>% select(sorted)

# Binding dfs
df <- rbind(df.2016, df.2017, df.2018, df.2019)

# Filtering data and removing rows with constant values
df <- df %>%
  filter(GeographicLevel == "Census Tract",
         !str_detect(GeoLocation, "POINT"),
         !is.na(TractFIPS)) %>%
  remove_constant()

# Creating final 500Cities dataframes
`500cities.unspread` <- df %>%
  select(TractFIPS, Year, MeasureId, Short_Question_Text, Measure, Data_Value) %>%
  distinct()
`500cities` <- df %>%
  select(TractFIPS, Year, MeasureId, Data_Value) %>%
  distinct() %>%
  spread(MeasureId, Data_Value)

# Removing unneeded dfs
rm(comparison, df.2016, df.2017, df.2018, df.2019)
```

#### Creating a reference dataframe for measures

```{r}
# Generating metadata file  of measures
temp1 <- df %>%
  select(MeasureId, Short_Question_Text, Measure) %>%
  distinct() %>%
  arrange(MeasureId)

# Getting the years of data for each measure
temp2 <- df %>%
  select(MeasureId, Year) %>%
  group_by(MeasureId) %>%
  distinct() %>%
  arrange(MeasureId, Year) %>%
  summarise(Years = paste(Year, collapse = ", "))

# Joining dataframes
metrics <- left_join(temp1, temp2)

# Removing unneeded dfs
rm(df, temp1, temp2)
```

<br>
<br>

### HOLC Grades Data

#### Combining HOLC data with spatial data

```{r}
# Loading HOLC data
df <- read_xlsx("MS_Data\ Summary.xlsx", sheet = 1)

# Selecting variables
df <- df %>%
  select(CensusTracts, holc_id, holc_grade) %>%
  dplyr::rename("TractFIPS" = CensusTracts,
                "HOLC_Id" = holc_id,
                "HOLC_Grade" = holc_grade)

# Loading spatial data to get percent of census tract land by grade
df2 <- read.dbf("Pct_HOLC.dbf")

# Selectng variables
df2 <- df2 %>%
  select(tract2010, holc_id, PERCENTAGE) %>%
  mutate(tract2010 = as.numeric(as.character(tract2010)),
         PERCENTAGE = round(PERCENTAGE / 100, 6)) %>%
  dplyr::rename("TractFIPS" = tract2010,
                "HOLC_Id" = holc_id,
                "HOLC_Pct" = PERCENTAGE)

# Joining HOLC and spatial data
df3 <- left_join(df, df2, by = c("TractFIPS", "HOLC_Id"))
```

#### Combining HOLC/spatial data with 500 cities data: Extra long format (for dynamic Shiny plot)

```{r}
# Joining HOLC/spatial data with unspread 500 cities data
data_xlong <- left_join(`500cities.unspread`, df3)
data_xlong <- data_xlong %>%
  select(TractFIPS, HOLC_Grade, HOLC_Pct, Year, MeasureId, Short_Question_Text, Measure, Data_Value) %>%
  distinct()

# Defining exposure variable - Only including grades that composed more than 30% of the census tract's land
data_xlong <- data_xlong %>%
  mutate(HOLC_Pct2 = ifelse(HOLC_Pct > 0.3, round(HOLC_Pct * 100, 2), NA),
         HOLC_Grade2 = ifelse(is.na(HOLC_Pct2), NA, HOLC_Grade),
         HOLC_Grade3 = factor(ifelse(is.na(HOLC_Grade2), 1,
                                     ifelse(HOLC_Grade2 == "A", 2,
                                            ifelse(HOLC_Grade2 == "B", 3,
                                                   ifelse(HOLC_Grade2 == "C", 4,
                                                          ifelse(HOLC_Grade2 == "D", 5, NA))))),
                              levels = 1:5,
                              labels = c("No grade", "A", "B", "C", "D"))) %>%
  select(-HOLC_Grade, -HOLC_Pct, -HOLC_Grade2) %>%
  dplyr::rename("HOLC_Pct" = HOLC_Pct2,
         "HOLC_Grade" = HOLC_Grade3) %>%
  select(TractFIPS, HOLC_Grade, HOLC_Pct, Year, MeasureId, Short_Question_Text, Measure, Data_Value) %>%
  distinct()
```


#### Combining HOLC/spatial data with 500 cities data: Long format

```{r}
# Joining HOLC/spatial data with 500 cities data
data_long <- left_join(`500cities`, df3)
data_long <- data_long %>%
  select(TractFIPS, Year, HOLC_Grade, HOLC_Pct, ACCESS2:TEETHLOST)

# Defining exposure variable - Only including grades that composed more than 30% of the census tract's land
data_long <- data_long %>%
  mutate(HOLC_Pct2 = ifelse(HOLC_Pct > 0.3, round(HOLC_Pct * 100, 2), NA),
         HOLC_Grade2 = ifelse(is.na(HOLC_Pct2), NA, HOLC_Grade),
         HOLC_Grade3 = factor(ifelse(is.na(HOLC_Grade2), 1,
                                     ifelse(HOLC_Grade2 == "A", 2,
                                            ifelse(HOLC_Grade2 == "B", 3,
                                                   ifelse(HOLC_Grade2 == "C", 4,
                                                          ifelse(HOLC_Grade2 == "D", 5, NA))))),
                              levels = 1:5,
                              labels = c("No grade", "A", "B", "C", "D"))) %>%
  select(-HOLC_Grade, -HOLC_Pct, -HOLC_Grade2) %>%
  dplyr::rename("HOLC_Pct" = HOLC_Pct2,
         "HOLC_Grade" = HOLC_Grade3) %>%
  select(TractFIPS, HOLC_Grade, HOLC_Pct, Year, ACCESS2:TEETHLOST) %>%
  distinct()
```

#### Combining HOLC/spatial data with 500 cities data: Short format

```{r}
# Compiling HOLC grades by census tract
df4 <- df3 %>%
  group_by(TractFIPS) %>%
  dplyr::summarize(HOLC_Grades = paste(unique(HOLC_Grade), collapse = ", "))

# Getting the percent of census tract land graded positive
df4.pos <- df3 %>%
  filter(HOLC_Grade == "A" | HOLC_Grade == "B") %>%
  group_by(TractFIPS) %>%
  dplyr::summarize(HOLC_Pos_Pct = round(sum(HOLC_Pct) * 100, 2))

# Getting the percent of census tract land graded negative
df4.neg <- df3 %>%
  filter(HOLC_Grade == "C" | HOLC_Grade == "D") %>%
  group_by(TractFIPS) %>%
  dplyr::summarize(HOLC_Neg_Pct = round(sum(HOLC_Pct) * 100, 2))

# Joining dfs
df5 <- left_join(df4, df4.pos)
df5 <- left_join(df5, df4.neg)

# Defining categorical exposure variables
df5 <- df5 %>%
  mutate(HOLC_Pos = as.numeric(str_detect(HOLC_Grades, "A|B")),
         HOLC_Neg = as.numeric(str_detect(HOLC_Grades, "C|D")),
         HOLC_Contains = factor(ifelse(HOLC_Pos == 1 & HOLC_Neg == 0, 1,
                                       ifelse(HOLC_Pos == 0 & HOLC_Neg == 1, 2,
                                              ifelse(HOLC_Pos == 1 & HOLC_Neg == 1, 3, NA))),
                                levels = 1:3,
                                labels = c("Positive", "Negative", "Both")),
         HOLC_Majority = factor(ifelse(HOLC_Pos_Pct > 0.5, 1,
                                       ifelse(HOLC_Neg_Pct > 0.5, 2, 3)),
                                levels = 1:3,
                                labels = c("Positive", "Negative", "No majority")),
         HOLC_Entirety = factor(ifelse(HOLC_Pos == 1 & HOLC_Neg == 0, 1,
                                       ifelse(HOLC_Pos == 0 & HOLC_Neg == 1, 2, 3)),
                                levels = 1:3,
                                labels = c("Positive", "Negative", "Both")))

# Joining HOLC/spatial data with 500 cities data
data_short <- left_join(`500cities`, df5)
data_short <- data_short %>%
  select(TractFIPS, HOLC_Grades:HOLC_Entirety, Year, ACCESS2:TEETHLOST) %>%
  distinct()

# Removing unneeded dfs
rm(df, df2, df3, df4, df4.pos, df4.neg, df5)
```

<br>
<br>

### Saving Data

```{r}
# Saving dfs as .rdata
write_rds(`500cities`, "data.rdata")
write_rds(metrics, "measures.rdata")
write_rds(data_xlong, "data_xlong.rdata")
write_rds(data_long, "data_long.rdata")
write_rds(data_short, "data_short.rdata")

# Saving dfs as .csv
write_csv(`500cities`, "data.csv")
write_csv(metrics, "measures.csv")
write_csv(data_xlong, "data_xlong.csv")
write_csv(data_long, "data_long.csv")
write_csv(data_short, "data_short.csv")
```

# Exploratory Analysis

### Loading Packages & Data

#### Loading packages

```{r}
# Creating a function for loading packages
ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)) 
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE)
}

# Loading packages
packages <- c("dplyr", "DT", "foreign", "ggplot2", "ggthemes", "Gmisc", "Hmisc", "janitor", "maditr",
              "readr", "readxl", "reshape", "rsconnect", "stringr", "tidyr")
ipak(packages)

# Setting the ggplot2 theme
theme_sarah <- theme_fivethirtyeight() %+replace%
  theme(
    panel.background  = element_rect(fill = "white", color = "#D2D2D2", linetype = 1),
    plot.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    legend.position = "right",
    axis.title = element_text(),
    panel.grid.major = element_line(size = 0.3, color = "#D2D2D2"),
    axis.ticks = element_line(size = 0.3, color = "#D2D2D2")
    )
theme_set(theme_sarah)

rsconnect::setAccountInfo(name='sarahbolongaita',
			  token='0F0ECD3C344E89ABC3B88F09DA328B1A',
			  secret='yXUEcH5Acp1oYieMQDb1xb/VdF18HaLBwwcPPfWO')
```

#### Loading data

```{r}
# Cleaning environment
rm(list = ls())

# Setting working directory

# Loading data
data_long <- readRDS("data_long.rdata")
data_short <- readRDS("data_short.rdata")
measures <- readRDS("measures.rdata")
```

<br>
<br>

### Exploratory Summary Statistics

```{r}
# Creating a wrapper for getting descriptive statistics
getTable1Stats <- function(x, digits = 1){
  getDescriptionStatsBy(x = x, 
                        by = data_short$HOLC_Contains,
                        digits = digits,
                        statistics = TRUE,
                        header_count = TRUE)
  
}

# Getting descriptive statistics 
mergeDesc("Uninsured" = getTable1Stats(data_short$ACCESS2),
          "High Blood Pressure" = getTable1Stats(data_short$BPHIGH),
          "Coronary Heart Disease" = getTable1Stats(data_short$CHD),
          "Annual Checkup" = getTable1Stats(data_short$CHECKUP),
          "Current Smoking" = getTable1Stats(data_short$CSMOKING),
          "Diabetes" = getTable1Stats(data_short$DIABETES),
          "High Cholesterol" = getTable1Stats(data_short$HIGHCHOL),
          "Physical Inactivity" = getTable1Stats(data_short$LPA),
          "Mental Health" = getTable1Stats(data_short$MHLTH),
          "Physical Health" = getTable1Stats(data_short$PHLTH),
          "Obesity" = getTable1Stats(data_short$OBESITY),
          "Sleep <7 hours" = getTable1Stats(data_short$SLEEP),
          "Stroke" = getTable1Stats(data_short$STROKE),
          htmlTable_args = list(align = "r"))
```

<br>
<br>

### Graphical Analysis

```{r}
# Box plot - ACCESS2
data_long %>%
  filter(!is.na(ACCESS2)) %>%
  ggplot(aes(x = HOLC_Grade, y = ACCESS2)) +
    geom_boxplot(fill = "darkolivegreen3", outlier.size = 0.7) +
    ggtitle("Uninsured in Jackson, 2013-2017") +
    xlab("HOLC grade history") + theme(axis.text.x = element_text(vjust = 0.5),
                     panel.grid.major.x = element_blank()) +
    ylab("Uninsured prevalence (%)") + scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
    theme(legend.position = "none")

# Box plot - OBESITY
data_long %>%
  filter(!is.na(OBESITY)) %>%
  ggplot(aes(x = HOLC_Grade, y = OBESITY)) +
    geom_boxplot(fill = "darkorange", outlier.size = 0.7) +
    ggtitle("Obesity in Jackson, 2013-2017") +
    xlab("HOLC grade history") + theme(axis.text.x = element_text(vjust = 0.5),
                     panel.grid.major.x = element_blank()) +
    ylab("Obesity prevalence (%)") + scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
    theme(legend.position = "none")

# Box plot - DIABETES
 data_long %>%
  filter(!is.na(DIABETES)) %>%
  ggplot(aes(x = HOLC_Grade, y = DIABETES)) +
    geom_boxplot(fill = "deepskyblue3", outlier.size = 0.7) +
    ggtitle("Diabetes in Jackson, 2013-2017") +
    xlab("HOLC grade history") + theme(axis.text.x = element_text(vjust = 0.5),
                     panel.grid.major.x = element_blank()) +
    ylab("Diabetes prevalence (%)") + scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
    theme(legend.position = "none")

# Box plot - MHLTH
data_long %>%
  filter(!is.na(MHLTH)) %>%
  ggplot(aes(x = HOLC_Grade, y = MHLTH)) +
    geom_boxplot(fill = "gold1", outlier.size = 0.7) +
    ggtitle("Poor mental health in Jackson, 2013-2017") +
    xlab("HOLC grade history") + theme(axis.text.x = element_text(vjust = 0.5),
                     panel.grid.major.x = element_blank()) +
    ylab("Poor mental health prevalence (%)") + scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
    theme(legend.position = "none")
```

<br>
<br>

### Regression Analysis

#### Linear regression - Exposure: HOLC grade

Aside from examining box plots and creating a dynamic shiny app to explore the data, we ran linear regressions on the four outcomes that we chose for analysis. The interpretation of each regression is below: 

Access to health care: For every step change between A-D in HOLC grades, 
Insurance Rates: 
Obesity:
Diabetes:
Mental Health: 

We note that the R^2 values for each association is very low. However, this is partially attributed to the fact that holc_grades are categorical not continuous data points. 

```{r}
mod.ACCESS2 <- lm(ACCESS2 ~ HOLC_Grade, data = data_long)
summary(mod.ACCESS2)

mod.OBESITY <- lm(OBESITY ~ HOLC_Grade, data = data_long)
summary(mod.OBESITY)

mod.DIABETES <- lm(DIABETES ~ HOLC_Grade, data = data_long)
summary(mod.DIABETES)

mod.MHLTH <- lm(MHLTH ~ HOLC_Grade, data = data_long)
summary(mod.MHLTH)
```


# Final Analysis 
Our exploration of data in this project show that redlining has a potential negative effect on a wide range of health measures. Because of our difficulty compiling the dataset around health outcomes and holc_grades we were unable to pull additional data regarding potential confounders such as race and area median income. Future analysis could involve these additional indicators which would allow us to assume causality about the relationship between redlining and health outcomes. 

An additional question that came up was the difference between areas graded ‘C’ and ‘D’. For the four indicators that we chose above: rate of uninsurance, obesity, diabetes, and poor mental health we noticed that often the spread of the data was much tighter for areas designated as ‘C’ versus those designated as ‘D’. In addition to this, areas designated as ‘C’ at times had a higher prevalence of negative health outcomes than areas designated as ‘D’. A potential additional question could look at the differences between areas labeled ‘C’ versus those labeled ‘D’. This question could be answered again by looking at demographic differences between these two census tracts. 

Finally, we only included data from 2013-2017. Additional analysis could look at a wider range of years to show the continued effect of redlining since the 1930s. 
